Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-01-20 doc/network.tex
trying to do some work

An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength (reproductive number, \RR).
Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \cite{anderson1991infectious, diekmann1990definition}.
However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can often be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
These two quantities are linked by generation-interval distributions \citep{wallinga2007generation, park2019practical}.

@@

In this study, we explore the temporal and spatial variation in the observed generation-interval distributions obtained from contact tracing.
We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
We discuss statistical approaches to estimating the intrinsic distributions and \RR\ from observed generation-interval distributions.

\swp{Might rename the sections later... but I think I like the current order...}
\section{Generation-interval distributions across time}

\subsection{Intrinsic kernel}

We begin by introducing an analytical framework.
Let $K_a(\tau)$ be an individual-level intrinsic kernel, i.e., the rate at which secondary infections are expected to be caused by an individual $a$ \citep{svensson2015influence}; \swp{Need to say something more about Sven; come back later}
\swp{TODO: read \cite{svensson2007note} and \cite{tomba2010some}}
an individual-level intrinsic kernel describes an intrinsic infectiousness of an infected individual.
\jd{Sven has earlier stuff I think that influenced me (and that probably has this framework.}
The population-level intrinsic kernel is given by integrating over individual variations:
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
\jd{Do we need to worry about assumptions for the rest of this \P? Once we have an aspect space we need to assume that individual properties are independent of risk of infection. Also, since that's wrong on our networks, do we have to figure out why we're approximately saved? I guess because of scaling?}
\swp{What do you mean by individual properties?}
The basic reproductive number -- average number of secondary cases caused by an average primary case in a fully susceptible population -- is defined as: 
\begin{equation}
\RR_0 = \int K(t).
\end{equation}
The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
\begin{equation}
g(\tau) = \frac{K(t)}{\RR_0}.
\end{equation}
Then, the renewal equation model can be used to model incidence over time:
\begin{equation}
i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds
\end{equation}
where $i(\tau)$ is incidence at time $\tau$ and $S(\tau)$ is the proportion of the population susceptible.
Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation \citep{lotka1907relation}:
\begin{equation}
\frac{1}{\RR} = \int g(\tau) \exp(-r \tau) d\tau.
\end{equation}

@@

The intrinsic generation-interval distribution provides a link between the speed (exponential growth rate, $r$) and the strength (reproductive number, $\RR$) of an epidemic via Euler-Lotka equation \citep{lotka1907relation} but assumes a homogeneously mixing population.
Instead, we can derive a relationship between $r$ and $\RR$ that accounts for the egocentric effect using the egocentric generation-interval distribution:
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\end{equation}
This is equivalent to the relationship between $r$ and $\RR$ that \cite{trapman2016inferring} derived under a network structure. \swp{Need to confirm; Maybe prove it in appendix or something...how?}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
Similarly, \Rhat\ will always be greater than the \emph{true} reproduction number \RR\ as it does not account for other spatial effects that can decrease the mean generation interval.

@@

We refer to the first method as the population-level method as it relies on the observed distribution aggregated across population.
As the observed generation-interval distribution is a weighted distribution of the effective distribution, the effective distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
Hence, the population-level method relies on the observed generation intervals and the exponential growth rate:
\begin{equation}
g(\tau) \propto \tsub{c}{exp}(\tau) \exp(r\tau).
\end{equation}
While the population-level method is simple and intuitive, it does not use all available information from contact tracing data as it does not take into account who infected whom.

@@

First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
During the initial exponential growth period (approximately between 9 and 20 weeks), both population-level and individual-level methods yield similar estimates of mean generation interval and reproductive number.
However, we observe opposite trends after the growth period (approximately after 20 weeks).
As the population-level method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
On the other hand, as the individual-level method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible is translated to underestimation of both mean generation interval and reproductive number.
we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).
Around the 14th week, reproductive number estimate based on the observed generation interval approximately matches the true reproduction number; this is due to initial overestimation of the initial growth rate and we generally expect reproduction to be underestimated. \swp{Maybe another figure in the supp}

@@

Finally, we simulate 100 epidemics on an empirical network [CITE] and compare the estimates of different methods.

>>>>

An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength (reproductive number, \RR).
Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
These two quantities are linked by generation-interval distributions \citep{wallinga2007generation, park2019practical}.

@@

In this study, we explore the temporal and spatial variation in the observed generation-interval distributions obtained from contact tracing.
We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
We discuss statistical approaches to estimating the effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.

\swp{I might rename the sections later but I think I like the current order.}
\section{Generation-interval distributions across time}

\subsection{Intrinsic generation-interval distribution}

Generation-interval distributions are often viewed from a population-level pespective, as a distribution describing \emph{expected} infection times of \emph{an average} infected individuals, but we can distinguish population-level distributions from individual-level distributions; 
making this distinction clear will be particularly useful when we discuss spatial components.
An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence}.

We begin by introducing an analytical framework.
Let $K_a(\tau)$ be an individual-level intrinsic kernel, i.e., the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence}; 
an individual-level intrinsic kernel describes an intrinsic infectiousness of an infected individual.
The population-level intrinsic kernel is given by integrating over individual variations:
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
\jd{Do we need to worry about assumptions for the rest of this \P? Once we have an aspect space we need to assume that individual properties are independent of risk of infection. Also, since that's wrong on our networks, do we have to figure out why we're approximately saved? I guess because of scaling?}
\swp{What do you mean by individual properties?}
Similar to the inidividual-level kernel, the population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} individual.

The basic reproductive number -- average number of secondary cases caused by an average primary case in a fully susceptible population -- is defined as: 
\begin{equation}
\RR_0 = \int K(t).
\end{equation}
The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
\begin{equation}
g(\tau) = \frac{K(t)}{\RR_0}.
\end{equation}
Then, the renewal equation model can be used to model incidence over time:
\begin{equation}
i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds
\end{equation}
where $i(\tau)$ is incidence at time $\tau$ and $S(\tau)$ is the proportion of the population susceptible.
Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation \citep{lotka1907relation}:
\begin{equation}
\frac{1}{\RR} = \int g(\tau) \exp(-r \tau) d\tau.
\end{equation}

@@

The intrinsic generation-interval distribution provides a link between the speed (exponential growth rate, $r$) and the strength (reproductive number, $\RR$) of an epidemic via Euler-Lotka equation \citep{lotka1907relation} but assumes a homogeneously mixing population.
Instead, we can derive a relationship between $r$ and $\RR$ that accounts for the egocentric effect using the egocentric generation-interval distribution:
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\end{equation}
This is equivalent to the relationship between $r$ and $\RR$ that \cite{trapman2016inferring} derived under a network structure. \swp{Need to confirm; Maybe prove it in appendix or something...how?}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
Similarly, \Rhat\ will always be greater than the \emph{true} reproductive number \RR\ as it does not account for other spatial effects that can decrease the mean generation interval.

@@

We refer to the first method as the population-level method as it relies on the observed distribution aggregated across population.
As the observed generation-interval distribution is a weighted distribution of the effective distribution, the effective distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
Hence, the population-level method relies on the observed generation intervals and the exponential growth rate \citep{tomba2010some}:
\begin{equation}
g(\tau) \propto \tsub{c}{exp}(\tau) \exp(r\tau).
\end{equation}
While the population-level method is simple and intuitive, it does not use all available information from contact tracing data as it does not take into account who infected whom.

@@

First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
During the initial exponential growth period (approximately between 9 and 20 weeks), both population-level and individual-level methods yield similar estimates of mean generation interval and reproductive number.
However, we observe opposite trends after the growth period (approximately after 20 weeks).
As the population-level method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
On the other hand, as the individual-level method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible is translated to underestimation of both mean generation interval and reproductive number.
we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).
Around the 14th week, reproductive number estimate based on the observed generation interval approximately matches the true reproductive number; this is due to initial overestimation of the initial growth rate and we generally expect reproductive to be underestimated. \swp{Maybe another figure in the supp}

@@

\begin{figure}
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{Fill this out.}
\label{fig:cmp}
\end{figure}

%% TODO: put SEIR in main text and show SE2IR in appendix for consistency
Finally, we simulate 100 epidemics with Ebola-liked parameters on an empirical network [CITE] and compare the estimates of reproductive number based on different methods with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects.
Direct estimates based on contact tracing data severely underestimates the empirical estimates.
Both population and individual corrections provide similar estimates to empirical reproductive numbers.
For smaller values of \RR, we expect the differences to become smaller.
